---
title: Data Preparation
subtitle: > 
  Data assembly and documentation with <font class="mrgc">mrgda</font>.
categories: 
- yspec
- mrgda
- data assembly
order: 300
---

# Introduction 

<hr/>

This page will demonstrate our data assembly workflow as we derive a data set
using a prepared data specification file and a source data set.

# Tools used
<hr/>

## MetrumRG Packages 

{{< var used.yspec >}}

{{< var used.mrgda >}}

{{< var used.lastdose >}}

## CRAN Packages

{{< var used.dplyr >}}

# Outline
<hr/>

A source data directory, coming as a collection of `.sas7bdat` files is provided. 
These files are found in the `data/source` folder. 

A data specification yaml file can be found at `data/derived/da-spec.yml`. 
A spec object will be created from this using the {{< var tools.yspec >}} package.

Below, we will assemble PK observations and dosing administration records, along
with baseline covariates to create a NONMEM ready data set.

The `mrgda` package will be used to:

- read the source data into R
- assign `ID` consistently to each subject
- write a NONMEM compliant csv file along with data assembly meta data

We will use the spec object and `yspec` package to:

- validate the derived dataset
- add labels to the derived dataset
- generate a data definition document in `pdf` format

# Setup

<hr/>

First, we will load the required packages, source data, and data specification file.

## Required packages

All packages will be installed from {{< var tools.mpn >}} via 
{{< var tools.pkgr >}}.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(mrgda)
library(yspec)
library(lastdose)
```

## Load in source data

The source data is read in using `read_src_dir()` from the `mrgda` package. It is
saved to the concise, descriptive variable name `src_list`.

```{r message=FALSE}
src_list <- read_src_dir(here("data", "source"))
```

`src_list` is a named list containing data.frames of each source domain. Below we can
see the contents of one of the domains, demographics (dm).

```{r}
head(src_list$dm)
```

## Load and examine the spec object

The spec file (`examp-da-spec.yml`) identifies the desired data columns 
for the data set we are deriving. We can load the spec file into an object in 
the R session using the `ys_load()` function from the `yspec` package.

```{r}
spec <- ys_load(here("data/derived/da-spec.yml"))
```

Each row in the object contains attributes about one column in the data set.
The first handful of rows can be viewed with the `head()` function.

```{r}
head(spec)
```

:::{.callout-note collapse="true" appearance="minimal"}
## Learn more about the spec object

Each row in the spec object can have:

- a name which must be $<=$ 8 characters long (by default)
  - continuous data items have units associated with them
  - discrete data items (like `SEQ` or `EVID`) can have valid levels listed in 
    the spec as well as decodes for those levels
- a "short" name that can be used in figure labels and tables
- a tex specific namespace so that the units can potentially be returned with LaTeX formatting
- a flags argument that can be used later to extract only covariate specific columns from the spec

:::

## Define output lists

During data assemblies, we work with two types of variables: subject level and
time-varying.

Subject level variables are those where there is only one unique value per 
subject. Sex, race and most demographic/baseline values are examples of these.

Time-varying variables are those that change with time. 
Dosing records, PK observations and time-varying lab values are examples of these.

To organize the variables into their respective types, we create output lists
below. All subject level variables are saved in the `derived$sl` list, while 
time-varying variables are in the `derived$tv` list.

```{r}
derived <- list()
derived$sl <- list()
derived$tv <- list()
```

# Demographics
<hr/>

This section will focus on deriving subject level variables from the
demographics source domain.

## Remove screen failures

First we filter the data to only the subjects we are interested in. Subjects who
do not meet the inclusion criteria for the study are marked as a `Screen Failure`.
We want to remove these subjects.

```{r}
dm0 <-
  src_list$dm %>%
  filter(ACTARM != "Screen Failure")
```

## Grab covariates of interest

According to the specification file, we need to derive `AGE` and `SEX`,
which can all be found in the `dm` domain.

`SEX` is a categorical variable, which means we need to assign each
category a numerical value according to the specification file. 

```{r}
spec$SEX
```

As shown above, males should be `SEX = 0` and females `SEX = 1`.

```{r}
dm1 <-
  dm0 %>%
  transmute(
    USUBJID,
    AGE,
    SEX = if_else(SEX == "F", 1, 0),
    ACTARM,
    STUDY = STUDYID,
    SUBJ = SUBJID
  )
```

## Save to subject level output list

With our variables of interest derived, we save `dm1` to the `derived$sl` list.

```{r}
derived$sl$dm <- dm1
```

# Labs
<hr/>

This section will focus on deriving subject level variables from the
labs source domain (`lb`).

From the specification file, we need to derive `ALT`, `AST`, `BILI`, `CREAT`
from the this domain.

## Filter to baseline values

The `LBBLFL` variable in the `lb` domain indicates which records are baseline
measurements. We can use this variable and the subjects we derived from
the `dm` domain to filter down to records and subjects of interest.

```{r}
lb0 <-
  src_list$lb %>%
  filter(LBBLFL == "Y") %>% 
  filter(USUBJID %in% derived$sl$dm$USUBJID) 
```

## Manipulate the format of the data

Each row in the `lb` domain has one lab test, meaning subjects have multiple 
rows in the data. We want to transform the data so that each subject only has
one row for multiple labs. 

`LBTESTCD` identifies the type of lab test and `LBSTRESN` contains the result. 
We want to update the data such that the `LBTESTCD` values are the names of the
columns, filled with the `LBSTRESN` values.

```{r}
lb1 <-
  lb0 %>%
  filter(LBTESTCD %in% c("ALT", "AST", "BILI", "CREAT")) %>%
  select(USUBJID, LBTESTCD, LBSTRESN) %>% 
  pivot_wider(names_from = "LBTESTCD", values_from = "LBSTRESN") 

lb2 <- lb1 %>% rename(SCR = CREAT)
```

## Save to subject level output list

Since `lb2` contains baseline labs, we save it to the `derived$sl` list.

```{r}
derived$sl$lb <- lb2
```

```{r, include=FALSE}
# Assemble baseline vitals ------------------------------------------------
vs0 <-
  src_list$vs %>%
  # Filter to only baseline measurements
  filter(VSBLFL == "Y") %>% 
  # Filter vs domain to only subjects in the dm output
  filter(USUBJID %in% derived$sl$dm$USUBJID) 

# Filter to only baseline weight measurements
vs1 <-
  vs0 %>%
  filter(VSTESTCD %in% c("WEIGHT")) %>%
  # Transform to create WEIGHT column
  pivot_wider(names_from = "VSTESTCD", values_from = "VSSTRESN") %>% 
  transmute(
    USUBJID,
    WT = WEIGHT
  )
# Save the subject level variables to the derived$sl list
derived$sl$vs <- vs1
```

# Dosing 
<hr/>

This section will focus on deriving time-varying dose administration records
from the `ex` domain. 

## Filter to desired treatment type

The `ex` domain contains dosing records for all treatments given to the subject.
We can use the `EXTRT` variable to filter the data to only treatments of interest.

```{r}
ex0 <-
  src_list$ex %>%
  filter(EXTRT %in% c("PLACEBO", "XANOMELINE"))
```

## Define variables of interest

According to the specification file, `EVID` and `DVID` need to be set to the
numeric decodes for dosing records. We check this below:

```{r}
spec$EVID
```

Dosing records also require an `AMT` variable representing the dose.
We also need to obtain the date/time of the dosing administration. This 
information can be found in the `EXSTDTC` variable.

```{r}
ex1 <-
  ex0 %>% 
  transmute(
    USUBJID,
    EVID = 1,
    DVID = 1,
    DOSE = EXDOSE,
    AMT = DOSE, 
    DATETIME = lubridate::ymd(EXSTDTC)
  )
```

## Save to time-varying output list

With all doing records captured, `ex1` can be saved to the `derived$tv` list.

```{r}
derived$tv$ex <- ex1
```

# PK observations
<hr/>

This section will focus on deriving time-varying pk observations from the `pc`
domain. 

## Filter to desired concentrations

The `PCTEST` variable contains a description of the concentration measured.
We want to filter to only the concentrations we are interested in.

```{r}
pc0 <-
  src_list$pc %>%
  filter(PCTEST == "XANOMELINE")
```

## Define variables of interest

Similar to dosing records, `EVID` and `DVID` need to be set to the
numeric decodes provided in the specification file. 

We are also interested in the `PCLLOQ` column. These values represent the
lower limit of quantification (LLOQ) of each concentration sample. If a
concentration is lower than the LLOQ, then the below limit of quantification
(BLQ) flag is set to 1. 

Additionally, the concentration values can be obtained from `PCSTRESN` and the
date/time of the record from `PCDTC`.

```{r}
pc1 <-
  pc0 %>%
  transmute(
    USUBJID,
    EVID = 0,
    DVID = 2,
    BLQ = if_else(grepl("<", PCORRES, fixed = TRUE), 1, 0),
    DV = if_else(BLQ == 0, PCSTRESN, NA_real_),
    LLOQ = PCLLOQ,
    DATETIME = lubridate::ymd_hms(PCDTC)
  )
```

## Save to time-varying output list

With all doing records captured, `pc1` can be saved to the `derived$tv` list.

```{r}
derived$tv$pc <- pc1
```

# Combine domains
<hr/>

This section will focus on combining the time-varying and subject level data 
we have assembled above. Once combined, final modifications will be made to 
match the data to the specification file.

## Bind and join derived data

The data we have saved to `derived$tv` will be used to generate rows in our 
combined data set. Meanwhile, the data within `derived$sl` will add columns.

First, we will combine the dataframes in `derived$sl`. Notice that we have 
one row per subject.

```{r}
baseline_variables <- reduce(derived$sl, full_join, by = "USUBJID")
head(baseline_variables)
```

Next, we will bind together the time-varying data and then join on 
`baseline_variables`. We sort by subject and date/time.

```{r}
nm0 <-
  bind_rows(derived$tv) %>%
  left_join(baseline_variables, by = "USUBJID") %>%
  arrange(
    USUBJID,
    DATETIME
  )
```

There are variables that need to be filled for every row within a subject
(`locf`).

```{r}
nm1 <-
  nm0 %>% 
  group_by(USUBJID) %>%
  tidyr::fill("DOSE", .direction = "downup") %>%
  ungroup()
```

## Derive TIME and TAD 

The {{< var tools.lastdose >}} package can be used to create `TIME` and `TAD` 
variables in the data set when the following columns are present:

- subject ID 
- record time 
- dose amount
- EVID

```{r}
nm2 <-
  nm1 %>%
  lastdose(include_tafd = TRUE, time_units = "hours") %>%
  mutate(
    TIME = TAFD
  )
```

## Assign ID variable

The {{< var tools.mrgda >}} package has a function `assign_id()` that can be 
used to derive the ID variable. It ensures each subject is assigned a consistent 
unique numerical value.

```{r, message=FALSE}
nm3 <-
  nm2 %>%
  assign_id(.subject_col = "USUBJID")
```

## Final modifications

To conclude the data assembly, we need to ensure the variables in our data set 
match those in the specification file. We can use `mrgmisc::pool()` to check
this:

```{r}
mrgmisc::pool(names(spec), names(nm3))
```

There are three more variables we need to add to the data: `C`, `NUM` and `MDV`. 
These are common NONMEM variables and can be derived as such. It's important
to wait to derive `NUM` until all other modifications have been made to the data.

```{r}
nm4 <-
  nm3 %>% 
  mutate(
    MDV = if_else(is.na(DV), 1, 0),
    C = ".",
    NUM = 1:n()
  )

derived$nm <- nm4 %>% select(names(spec))
```

The final derivation step is to save our final data set to `derived$nm`.

## Verify derived data with yspec

Before submitting our derived data set, we can leverage the spec object to 
prepare and verify our work.

`yspec` can compare the candidate data set with what is in the specification. 
The `ys_check()` function will compare the data set and spec object. It will
output a message saying the result of the comparison.

```{r, message = TRUE, warning}
ys_check(derived$nm, spec)
```

Everything looks to be in order! 

:::{.callout-note collapse="true" appearance="minimal"}
## Learn more about <font class="mrgc">ys_check</font>

The call to `ys_check()` does a limited check of the data against 
the spec, including

1. all the columns are present and in the correct order
2. levels of discrete data columns match up between the spec and 
the data
3. the range for all continuous data items is within the range 
specified in the spec

Note also that certain data requirements are also enforced on load, 
for example

1. all column names contain 8 or fewer characters
2. the `short` field in the spec contains 40 or fewer characters for 
all columns
3. the `label` field in the spec contains 40 or fewer characters for 
all columns
:::

## Output data to csv

Since the data and spec object match, our final step is to write the data to
`data/derived/examp-da.csv`. We can use the `write_derived()` function from 
`mrgda`.

```{r, eval = TRUE}
write_derived(
  .data = derived$nm,
  .spec = spec,
  .file = "data/derived/examp-da.csv"
)
```

You can view our finished product in the `data/derived` folder.

# Documenting derived data
<hr/>

This section will provide an example for creating a define document and SAS 
xport file.

## Write a data definitions document

We create a define document for `examp-da.csv` in the `data/derived` folder. 

```{r, eval = TRUE}
ys_document(
  spec, 
  type = "regulatory", 
  output_dir = here("data/derived"), 
  output_file = "examp-da.pdf", 
  build_dir = definetemplate(), 
  author = "Kyle Baron"
)
```

When rendering the define document, the `regulatory` type is selected, which is 
set up to follow the format requested for pharmacometrics submissions. The 
`definetemplate()` format puts a more polished style on the output, but it 
can be omitted or updated.

We can look at our define document here `data/derived/examp-da.pdf`.

# Other resources

<hr />

The following script from the {{< var expo_repo.main >}} is discussed on this page. If you're interested running this code, visit the {{< var pages.about_the_repo >}} page first.

- Runnable version of the examples on this page: {{< var expo_repo.data_prep >}}

Further uses of the package {{< var tools.yspec >}} can be seen in the
exploratory data analysis (EDA) pages.

To advance to the EDA content, go here: 

- {{< var pages.eda_figures >}}
- {{< var pages.eda_tables >}}
